---
title: "HW04"
output: github_document
editor_options: 
  chunk_output_type: console
---

# Homework 04: Tidy data and joins

Load packages

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
```

## Data Reshaping Prompts (and relationship to aggregation)

### Activity 2

Make a tibble with one row per year and columns for life expectancy for two or more countries.

```{r}
gap_subset <- gapminder %>%
  filter(country %in% c("Iceland", "Japan", "Kenya")) %>%
  select(country, year, lifeExp)

gap_subset %>%
  knitr::kable()
```

Here we can see the gapminder subset with 3 countries in a long format (with year and lifeExp).

```{r}
gap_subset %>%
  spread(key = "country", value = "lifeExp") %>%
  knitr::kable()
```

After `spread`, we can see that the columns are the lifeExp for the 3 countries and each row is a year.

```{r}
gap_subset %>%
  ggplot(aes(x = year, y = lifeExp, color = country)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  ylab("Life Expectancy")
```

Now we have a scatter plot with line to visualize. Note, I have changed the theme and the axis label on the Y.

### Activity 3

Compute some measure of life expectancy (mean? median? min? max?) for all possible combinations of continent and year. Reshape that to have one row per year and one variable for each continent. Or the other way around: one row per continent and one variable per year.

```{r}
med_lifeEx <- gapminder %>%
  group_by(continent, year) %>%
  summarise(med = median(lifeExp))

med_lifeEx %>%
  knitr::kable()
```

Here we can see the median lifeExp for each continent each year. It is in a long format.

```{r}
med_lifeEx %>%
  spread(key = "continent", value = "med") %>%
  knitr::kable()
```

Here we can see it in a `spread` form with one column per continent.

```{r}
med_lifeEx %>%
  spread(key = "continent", value = "med") %>%
  gather(key = "continent", value = "med", -"year") %>%
  knitr::kable()
```

We can see that `gather` and reverts `spread` back to the original form.

```{r}
med_lifeEx %>%
    ggplot(aes(x = year, y = med, color = continent)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  ylab("Median Life Expectancy")
```

Now we have a scatter plot with line to visualize. Note, I have changed the theme and the axis label on the Y. Additionally, the gathered format is easier for plotting, but the spread format is nicer for viewing in a table.

## Join Prompts (join, merge, look up)

### Activity 1

Create a second data frame, complementary to Gapminder. Join this with (part of) Gapminder using a dplyr join function and make some observations about the process and result. Explore the different types of joins. Examples of a second data frame you could build:
One row per country, a country variable and one or more variables with extra info, such as language spoken, NATO membership, national animal, or capitol city.

First look at the gapminder col names.

```{r}
colnames(gapminder)
```

Read in a text file with country capital cities.

```{r}
country_cap <- read_delim("Country_capitals.txt", delim = "$", col_names = TRUE)

head(country_cap, 20) %>%
  knitr::kable()
```

Here, I show that we can use `read_delim` and I specifically formatted the file with "$" as the delimeter. This file has countries and capitals (shown with `head` of 20). I call this `country_cap`.

```{r}
gapminder %>%
  filter(year == 1952) %>%
  left_join(country_cap, by = "country") %>%
  knitr::kable()
```

I first filter `gapminder` for data in 1952. Then I `left_join` with `country_cap`. I can see the `gapminder` entries, but now with a new column of `captial`. We can see some `NA`. This means that the country name was not in `country_cap`.

```{r}
gapminder %>%
  filter(year == 1952) %>%
  right_join(country_cap, by = "country") %>%
  knitr::kable()
```

With `right_join`, we pick all the entries in `country_cap` and then add the columns that were also present in `gapminder`. Once again, note the `NA`.

```{r}
gapminder %>%
  filter(year == 1952) %>%
  inner_join(country_cap, by = "country") %>%
  knitr::kable()
```

Note: this time, there are no `NA`. With `inner_join`, we only have the intersected country name data shown (those in common).

```{r}
gapminder %>%
  filter(year == 1952) %>%
  semi_join(country_cap, by = "country") %>%
  knitr::kable()
```

With `semi_join`, we see all the rows in `gapminder` that have a country present in `country_cap`. So which countries are missing?

```{r}
gapminder %>%
  filter(year == 1952) %>%
  anti_join(country_cap, by = "country") %>%
  knitr::kable()
```

These rows belong to countries found in `gapminder`, but not `country_cap`.

```{r}
anti_join(country_cap, gapminder, by = "country") %>%
  knitr::kable()
```

Here, this `anti_join` will now show us countries present in `country_cap`, but not `gapminder`. Note that this differs from before! Order matters!

These 2 `anti-join`s above show us that in data analysis, we should see what things may not match. In `gapminder`, we can see countries like "Congo, Dem. Rep." and "United States", Meanwhile, in `country_cap`, we have "Congo, Democratic Republic of the" and "United States of America". In fact, these are the same countries, but the spelling of the entries are different.

Related to my own research, I might have 2 gene sets. And if I perform joins, I may have errors if a gene is referred to by 2 different names.

## Activity 2

Create your own cheatsheet patterned after Jenny’s but focused on something you care about more than comics! 

Since I demonstrated various join functions above, I'm going to pick some set operations now.

First I'll conduct 2 surveys about people and their favourite food.

```{r}
survey1 <- tibble(name = c("Alex", "Blake", "Charlie", "Diane"),
                   food = c("Apple", "Bacon", "Cheese", "Donuts"))

survey2 <- tibble(name = c("Blake", "Diane", "Eliza", "Frank"),
                   food = c("Bacon", "Donuts", "Eggs", "Freezies"))

survey1 %>%
  knitr::kable()

survey2 %>%
  knitr::kable()

```

Let's try `bind_rows`.

```{r}
bind_rows(survey1, survey2) %>%
  knitr::kable()
```

We can see that the 2 surveys get appended to each other. But there are repeats?

```{r}
intersect(survey1, survey2) %>%
  knitr::kable()
```

Here is the `intersect`. We can see that the 2 surveys had overlap in who they interviewed. If we just want all of results, but without repeats (seen in row binds), we can use `union`.

```{r}
union(survey1, survey2) %>%
  arrange(name) %>%
  knitr::kable()
```

Here is now a complete list of both surveys (using `arrange` to maintain alphabetical order).

Who was unique to survey 1?

```{r}
setdiff(survey1, survey2) %>%
  knitr::kable()
```

And who was unique to survey 2?

```{r}
setdiff(survey2, survey1) %>%
  knitr::kable()
```

`setdiff` lets us answer these questions. Note that the order matters!

Now let's add a survey 3 to show a few more things. This has people's favourite movies.

```{r}
survey3 <- tibble(name = c("Alex", "Blake", "Charlie", "Diane"),
                  Movie = c("Avengers", "Bambi", "Coco", "Dirty Dancing"))

survey3 %>%
  knitr::kable()
```

Let's `bind_cols`.

```{r}
bind_cols(survey1, survey3) %>%
  knitr::kable()
```

Note: that the columns are bound, but this is not ideal. The name column is repeated.

```{r}
bind_cols(survey1, survey3 %>% select(Movie)) %>%
   knitr::kable()
```

We could try this solution above.

```{r}
inner_join(survey1, survey3) %>%
   knitr::kable()
```

Better yet is using a join. Note: that `dyplr` automatically joins with "name".